---
title: "Define MSBB Variables"
description: |
 This script first matches MSBB expression data to the clinical data, then defines variables in the covariate data.
output:
  distill::distill_article:
    toc: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

# Dependencies

Load requisite packages.

```{r load-packages}

library(data.table)

```


# Define Paths

Define brain regions for analysis and paths to normalized data files (i.e., Trimmed Mean of M-Values or TMM matrices).

```{r read-data}

brain_regions = c("IFG", "STG", "PHG", "FP")
data_files = list(
  IFG = "../Data/AMP-AD_MSBB_MSSM_BM_44.normalized.sex_race_age_RIN_PMI_exonicRate_rRnaRate_batch_adj.tsv",
  STG = "../Data/AMP-AD_MSBB_MSSM_BM_22.normalized.sex_race_age_RIN_PMI_exonicRate_rRnaRate_batch_adj.tsv",
  PHG = "../Data/AMP-AD_MSBB_MSSM_BM_36.normalized.sex_race_age_RIN_PMI_exonicRate_rRnaRate_batch_adj.tsv",
  FP = "../Data/AMP-AD_MSBB_MSSM_BM_10.normalized.sex_race_age_RIN_PMI_exonicRate_rRnaRate_batch_adj.tsv"
)

```


# Analyze Brain Regions

Iterate over previously specified brain regions to process each region. **All code after this chunk should be encapsulated in a `for` loop.** To appropriately document the code, pseudocode of the `for` loop is shown below, while the contents of the `for` loop are described in subsequent chunks.

```{r analyze-regions}

for(brain_region in brain_regions) {
  
  # insert following chunks here
  
}

```

For the purposes of illustration, we define the `brain_region` of interest as the inferior frontal fygrus (IFG).

```{r define-region}

brain_region = "IFG"

```


# Process Expression Data

Process expression data and remove unnecessary headers from sample IDs (e.g., `S109B355.BM_10_798` becomes `BM_10_798`).

```{r process-expression}

# process expression data
cat("processing", brain_region, "...\n")
data_file = unlist(data_files[brain_region])
msbb = suppressWarnings({fread(data_file, sep = "\t")})
gene_ids = msbb$V1
msbb = as.data.frame(msbb[, -1, with = F])
rownames(msbb) = gene_ids

# remove unnecessary headers
colnames(msbb) = gsub("^.*\\.", "", colnames(msbb))

```

Extract subject IDs in the same order as the sample IDs, then check for samples without an ID match.

```{r extract-subjects}

# extract subject IDs
key_file =  "../Data/MSBB_RNAseq_covariates_November2018Update.csv"
key_map = fread(key_file)
key_map = unique(key_map[, .(sampleIdentifier, individualIdentifier)])

# check for samples without an ID match
cat("any samples don't have id match? ")
cat(any(!colnames(msbb) %in% key_map$sampleIdentifier), "\n")
sub_key_map = key_map[data.table(sampleIdentifier = colnames(msbb)), on = .(sampleIdentifier)]
all(sub_key_map$sampleIdentifier == colnames(msbb))
colnames(msbb) = sub_key_map$individualIdentifier


```

Check for duplicated samples.

```{r check-duplicates}

# any duplicated samples?
if(any(duplicated(colnames(msbb)))){
  cat("remove duplicate samples..\n")
  msbb = msbb[, !duplicated(colnames(msbb))]
}

```


# Process Covariate Data

Note that subject #1009 has an invalid Braak score of 9.

```{r process-covariate}

cov_file = "../Data/MSBB_clinical.csv"
cov = fread(cov_file)
setnames(cov, c("NP.1", "bbscore"), c("CERAD", "Braak"))
cov[Braak > 6, Braak := NA]

```

Extract clinical data in the same order as in the data.

```{r extract-clinical}

cov = cov[data.table(individualIdentifier = colnames(msbb)), on = .(individualIdentifier)]
all(cov$individualIdentifier == colnames(msbb))

```


# Define Clinical Variables

## Braak

<p>Group levels into three categories:</p>
<ul>
  <li>`1` = `Normal/I/II`</li>
  <li>`2` = `III/IV`</li>
  <li>`3` = `V/VI`</li>
</ul>

```{r braak}

cov[, B := as.factor(cov$Braak)]
levels(cov$B) = c(1,1,1,2,2,3,3)
cov[, B := as.factor(as.character(B))]
table(cov$B, cov$Braak)

# double-check
table(cov$C, cov$Braak)

```

## CERAD

<p>Old MSBB levels:</p>
<ul>
  <li>`1` = `Normal`</li>
  <li>`2` = `Definite`</li>
  <li>`3` = `Probable`</li>
  <li>`4` = `Possible`</li>
</ul>

<p>Define new levels:</p>
<ul>
  <li>`1` = `Normal`</li>
  <li>`2` = `Possible`</li>
  <li>`3` = `Probable`</li>
  <li>`4` = `Definite`</li>
</ul>

```{r cerad}

# refactor
cov[, C := as.factor(CERAD)]
levels(cov$C) = c(0,3,2,1)
cov[, C := as.factor(as.character(C))]

# double-check
table(cov$C, cov$CERAD)

```

## Raw APOE

<p>Define three categories:</p>
<ul>
  <li>`1` = `E2/E2` and `E3/E3`</li>
  <li>`2` = `E3/E4`, `E2/E4`, and `E4/E4`</li>
</ul>

`NA` values will be treated as `E3/E3`.

```{r raw-apoe}

# define variable
cov$RawAPOE = paste0(cov$Apo1, cov$Apo2)
cov$RawAPOE = gsub("NANA", NA, cov$RawAPOE)

# treat NA values as E3/E3
cov[is.na(RawAPOE), RawAPOE := "33"]

```

## APOE

<p>Group levels into three categories:</p>
<ul>
  <li>`E2` = `E2/E2` and `E2/E3`</li>
  <li>`E3` = `E3/E3`</li>
  <li>`E4` = `E2/E4`, `E3/E4`, and `E4/E4`</li>
</ul>

```{r apoe}

# refactor
levels(cov$E) = c("E2", "E2", "E4", "E3", "E4", "E4")
cov[, E := factor(as.character(E), levels = c("E3", "E2", "E4"))]

# double-check
table(cov$E, cov$RawAPOE)

```

## E4 Count

<p>Group levels into three categories:</p>
<ul>
  <li>`0` = `E2/E2`, `E3/E3`, and `E2/E3`</li>
  <li>`1` = `E2/E4` and `E3/E4`</li>
  <li>`2` = `E4/E4`</li>
</ul>

```{r e4-count}

# refactor
cov[, E4num := as.factor(RawAPOE)]
levels(cov$E4num) = c(0, 0, 1, 0, 1, 2)
cov[, E4num := as.numeric(as.character(E4num))]

# double-check
table(cov$E4num, cov$RawAPOE)

```

## E2 Count

<p>Group levels into three categories:</p>
<ul>
  <li>`0` = `E4/E4`, `E3/E3`, and `E3/E4`</li>
  <li>`1` = `E2/E4` and `E2/E3`</li>
  <li>`2` = `E2/E2`</li>
</ul>

```{r e2-count}

# refactor
cov[, E2num := as.factor(RawAPOE)]
levels(cov$E2num) = c(2, 1, 1, 0, 0, 0)
cov[, E2num := as.numeric(as.character(E2num))]

# double-check
table(cov$E2num, cov$RawAPOE)

```

## Age at Death

Clean age at death variable.

```{r age-death}

# clean age at death
cov[, age_at_death := gsub("90\\+", "91", AOD)]
cov[, age_at_death := as.numeric(age_at_death)]

```


# Save Results

Choose levels and contrasts, then save results.

```{r save-results}

# save results
expSet = list(mData = msbb, cov = cov)
save(expSet, file = paste0("../Data/MSBB-", brain_region,"-24.Rdata"))

```