---
title: "ROSMAP Astrocyte Analysis"
description: |
 This script performs analysis of ROSMAP astrocyte data.
output:
  distill::distill_article:
    toc: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

# Dependencies

Load requisite packages.

Load requisite packages. Note that package [`cttobin/ggthemr`](https://github.com/cttobin/ggthemr) hosted on GitHub is used to provide `ggplot2` themes. This package can be downloaded via `devtools::install_github("cttobin/ggthemr")`.

```{r load-packages}

rm(list = ls())
require(ggVennDiagram)
require(limma)
require(latex2exp)
require(ComplexHeatmap)
require(data.table)
require(pheatmap)
require(ggplot2)
require(gridExtra)
require(ggthemr)
require(SNFtool)
require(circlize)
require(patchwork)
require(cowplot)
require(ggpubr)

ds = "ROSMAP"
dtype = "log2FPKM"
load("../Data/ROSMAP-24-adj-low.expr.genes.removed.Rdata")

source("models.R")
source("spectral-clustering.R")

# load expression data 
mData = expSet$mData

```


# Select Astrocyte Genes

First, select only astrocyte genes (n = 405). Of the microglial genes, select only those that exist in ROSMAP (n = 397).

```{r select-astrocyte}

celltype = "astro"

# select only astrocyte genes
genes = readLines("../Data/Astrocyte Genes.txt")

# select astrocyte genes that exist in ROSMAP
annot = fread("../Data/ENSEMBL GRCh38.p7.csv")
annot = unique(annot[GeneSymbol %in% genes,])
ensembl_ids = intersect(rownames(mData), annot$EnsemblID)

annot = unique(annot[EnsemblID %in% ensembl_ids,])
# remove duplicated genes
annot = annot[!duplicated(GeneSymbol), ]
ensembl_ids = annot$EnsemblID

any(duplicated(ensembl_ids)) # no duplicated ids
any(is.na(ensembl_ids))      # no NA
length(ensembl_ids)

```


# Pre-Processing

```{r pre-processing}

# select interesting genes
mData = mData[annot$EnsemblID, ]
all(rownames(mData) == annot$EnsemblID)
rownames(mData) = annot$GeneSymbol

# set metadata
cov = expSet$cov

```


# Run Models
* Model 1: `exp ~ APOE + CERAD | exp ~ APOE + Braak`
* Model 2 baseline: `C0E4-C0E3, C0E2-C0E3 | B1E4-B1E3, B1E2-B1E3`
* Model 3: `C3E4-C3E3, C3E2-C3E3 | B3E4-B3E3, B3E2-B3E3`

## Model 1

Run Model 1, which is `exp ~ APOE + CERAD | exp ~ APOE + Braak`. 

```{r model-1}

# run Model 1
tT_CERAD = run_model1(mData, cov$C, cov$E, annot)
tT_Braak = run_model1(mData, cov$B, cov$E, annot)

length(tT_CERAD[E4vsE3_P.Value < 0.05 & E4vsE3_logFC > 0,]$GeneSymbol)  
length(tT_CERAD[E4vsE3_P.Value < 0.05 & E4vsE3_logFC < 0,]$GeneSymbol)

length(tT_CERAD[E2vsE3_P.Value < 0.05 & E2vsE3_logFC > 0,]$GeneSymbol)  
length(tT_CERAD[E2vsE3_P.Value < 0.05 & E2vsE3_logFC < 0,]$GeneSymbol)  

length(tT_Braak[E4vsE3_P.Value < 0.05 & E4vsE3_logFC > 0,]$GeneSymbol)  
length(tT_Braak[E4vsE3_P.Value < 0.05 & E4vsE3_logFC < 0,]$GeneSymbol)  

length(tT_Braak[E2vsE3_P.Value < 0.05 & E2vsE3_logFC > 0,]$GeneSymbol)  
length(tT_Braak[E2vsE3_P.Value < 0.05 & E2vsE3_logFC < 0,]$GeneSymbol) 

TableS2.astro = tT_CERAD[, .(EnsemblID, GeneSymbol, E4vsE3_logFC, E4vsE3_P.Value, E2vsE3_logFC, E2vsE3_P.Value)]

```

## Model 2

Run Model 2 baseline, which is `C0E4-C0E3, C0E2-C0E3 | B1E4-B1E3, B1E2-B1E3`. 

```{r model-2}

# run Model 2
tT2_CERAD = run_model2("CERAD", mData, cov$C, cov$E, annot) 
tT2_Braak = run_model2("Braak", mData, cov$B, cov$E, annot)

length(tT2_CERAD[E4vsE3_P.Value < 0.05 & E4vsE3_logFC > 0,]$GeneSymbol) 
length(tT2_CERAD[E4vsE3_P.Value < 0.05 & E4vsE3_logFC < 0,]$GeneSymbol)
length(tT2_Braak[E4vsE3_P.Value < 0.05 & E4vsE3_logFC > 0,]$GeneSymbol)
length(tT2_Braak[E4vsE3_P.Value < 0.05 & E4vsE3_logFC < 0,]$GeneSymbol)

```

## Model 3

Run Model 3, which is `C3E4-C3E3, C3E2-C3E3 | B3E4-B3E3, B3E2-B3E3`.

```{r model-3}

# run Model 3
tT3_CERAD = run_model3("CERAD", mData, cov$C, cov$E, annot) 
tT3_Braak = run_model3("Braak", mData, cov$B, cov$E, annot)

length(tT3_CERAD[E4vsE3_P.Value < 0.05 & E4vsE3_logFC > 0,]$GeneSymbol) 
length(tT3_CERAD[E4vsE3_P.Value < 0.05 & E4vsE3_logFC < 0,]$GeneSymbol)
length(tT3_Braak[E4vsE3_P.Value < 0.05 & E4vsE3_logFC > 0,]$GeneSymbol)
length(tT3_Braak[E4vsE3_P.Value < 0.05 & E4vsE3_logFC < 0,]$GeneSymbol)

```


# Z-Score Analysis

Compute z-scores.

```{r compute-z}

z.C0 = to_z_score("C", 0, mData, cov)
z.C3 = to_z_score("C", 3, mData, cov)
z.B1 = to_z_score("B", 1, mData, cov)
z.B3 = to_z_score("B", 3, mData, cov)

```

Calculate average z-scores.

```{r average-z}

avez.C0 = to_ave_z(z.C0)
avez.C3 = to_ave_z(z.C3)
avez.B1 = to_ave_z(z.B1)
avez.B3 = to_ave_z(z.B3)

```


# Spectral Clustering

## C0
* Number of Clusters: 5
* Index of Cluster of Interest: 4 and 1
* Number of C0 Genes: 69 in Cluster #4, 75 in Cluster #1

Note that Cluster #4 above was originally Cluster #3; they are switched for convenience.

```{r c0-clustering}

# spectral clustering
c = 4; k = 5
clust.C0 = spectral_clustering(avez.C0, k)
clust.C0$clustA = as.factor(clust.C0$clustA)

# switch clusters 3 and 4 for convenience
levels(clust.C0$clustA) = c("1", "2", "4", "3", "5")
clust.C0$clustA = as.numeric(as.character(clust.C0$clustA))

# visualization
annot_row = data.frame(cluster = as.character(clust.C0$clustA))
rownames(annot_row) = rownames(clust.C0$zMtx)
pheatmap(clust.C0$zMtx[order(clust.C0$clustA),],
         annotation_row = annot_row[order(clust.C0$clustA), , drop = F],
         cluster_rows = F, cluster_cols = F,
         show_rownames = F)

# interested genes
C0_c4.genes = sort(rownames(clust.C0$zMtx)[clust.C0$clustA == c])
length(C0_c4.genes)
C0_c1.genes = sort(rownames(clust.C0$zMtx)[clust.C0$clustA == 1])
length(C0_c1.genes)

```

## C3
* Number of Clusters: 5
* Index of Cluster of Interest: 5
* Number of C3 Genes: 68

```{r c3-clustering}

# spectral clustering
c = 5; k = 5
clust.C3 = spectral_clustering(avez.C3, k)

# visualization
annot_row = data.frame(cluster = as.character(clust.C3$clustA))
rownames(annot_row) = rownames(clust.C3$zMtx)
pheatmap(clust.C3$zMtx[order(clust.C3$clustA),],
         annotation_row = annot_row[order(clust.C3$clustA), , drop = F],
         cluster_rows = F, cluster_cols = F,
         show_rownames = F)

# interested genes
C3.genes = sort(rownames(clust.C3$zMtx)[clust.C3$clustA == c])
length(C3.genes)

```

## B1
* Number of Clusters: 5
* Index of Cluster of Interest: 6
* Number of B1 Genes: 70

```{r b1-clustering}

# spectral clustering
k = 5; c = 5
clust.B1 = spectral_clustering(avez.B1, k)

# visualization
annot_row = data.frame(cluster = as.character(clust.B1$clustA))
rownames(annot_row) = rownames(clust.B1$zMtx)
pheatmap(clust.B1$zMtx[order(clust.B1$clustA),],
         annotation_row = annot_row[order(clust.B1$clustA), , drop = F],
         cluster_rows = F, cluster_cols = F,
         show_rownames = F)

# interested genes
B1.genes = sort(rownames(clust.B1$zMtx)[clust.B1$clustA == c])
length(B1.genes)

```

## B3
* Number of Clusters: 5
* Index of Cluster of Interest: 1
* Number of B3 Genes: 76

```{r b4-clustering}

# spectral clustering
k = 5; c = 1
clust.B3 = spectral_clustering(avez.B3, k)

# visualization
annot_row = data.frame(cluster = as.character(clust.B3$clustA))
rownames(annot_row) = rownames(clust.B3$zMtx)
pheatmap(clust.B3$zMtx[order(clust.B3$clustA),],
         annotation_row = annot_row[order(clust.B3$clustA), , drop = F],
         cluster_rows = F, cluster_cols = F,
         show_rownames = F)

# interested genes
B3.genes = sort(rownames(clust.B3$zMtx)[clust.B3$clustA == c])
length(B3.genes)

```

## Overlap

```{r hypergeometric-tests}

# hypergeometric tests
phyper_test(C0_c4.genes, B1.genes)
phyper_test(C3.genes, B3.genes)
phyper_test(C0_c4.genes, C3.genes)
phyper_test(B1.genes, B3.genes)

```


# Save Results

Save results to Excel files.

```{r save-results}

# CERAD genes
xlsx::write.xlsx(C0_c4.genes, 
    "../results/ROSMAP_astrocyte_new_cluster_remove.low.expressed.genes.xlsx", 
                 sheetName = "C0_clust4(k=5)", append = TRUE, row.names = F, 
                 col.names = F)
xlsx::write.xlsx(C0_c1.genes, 
    "../results/ROSMAP_astrocyte_new_cluster_remove.low.expressed.genes.xlsx", 
                 sheetName = "C0_clust1(k=5)", append = TRUE, row.names = F, 
                 col.names = F)

xlsx::write.xlsx(C3.genes, 
    "../results/ROSMAP_astrocyte_new_cluster_remove.low.expressed.genes.xlsx", 
                 sheetName = "C3_clust1(k=5)", append = TRUE, row.names = F, 
                 col.names = F)

length(intersect(C0_c4.genes, C3.genes)) 
xlsx::write.xlsx(intersect(C0_c4.genes, C3.genes), 
      "../results/ROSMAP_astrocyte_new_cluster_remove.low.expressed.genes.xlsx", 
      sheetName = "intersect C0_cluster3&C4", append = TRUE, row.names = F, col.names = F)

# Braak genes
xlsx::write.xlsx(B1.genes, 
    "../results/ROSMAP_astrocyte_new_cluster_remove.low.expressed.genes.xlsx", 
                 sheetName = "B1_clust5(k=5)", append = TRUE, row.names = F, 
                 col.names = F)

xlsx::write.xlsx(B3.genes, 
    "../results/ROSMAP_astrocyte_new_cluster_remove.low.expressed.genes.xlsx", 
                 sheetName = "B3_clust1(k=5)", append = TRUE, row.names = F, 
                 col.names = F)

length(intersect(B1.genes, B3.genes)) 
xlsx::write.xlsx(intersect(B1.genes, B3.genes), 
      "../results/ROSMAP_astrocyte_new_cluster_remove.low.expressed.genes.xlsx", 
      sheetName = "intersect B1&B3", append = TRUE, row.names = F, col.names = F)

```

# Venn Diagram

Prepare Venn diagram.

```{r venn-diagram}

rosmap_genes = list(
  `C0 clust4` = C0_c4.genes,
  C3 = C3.genes,
  B1 = B1.genes,
  B3 = B3.genes
)
ggVennDiagram(rosmap_genes, label_alpha=0)
ggsave("../results/venn_rosmap_astrocyte.pdf")

ggVennDiagram(rosmap_genes[c("C0 clust4", "C3")], label_alpha=0)
ggsave("../results/venn_rosmap_astrocyte_C0_c4&C3.pdf")
ggVennDiagram(rosmap_genes[c("B1", "B3")], label_alpha=0)
ggsave("../results/venn_rosmap_astrocyte_B1&B3.pdf")

```


# Statistical Testing

Statistical testing of astrocyte-APOE genes across *APOE* genotypes. Average of Z-scores for interested genes for each individual as outcome, with *APOE* genotype as a covariate. Use *APOE* `E3/E3` as the baseline.

```{r statistical-testing}

# C0 test
cov_C0 = cov[C == 0,]
all(cov_C0$individualIdentifier == colnames(z.C0$z))
test_avez(z.C0$z, C0_c4.genes, cov_C0)

# C3 test
cov_C3 = cov[C == 3,]
all(cov_C3$individualIdentifier == colnames(z.C3$z))
test_avez(z.C3$z, C3.genes, cov_C3)

# B1 test
cov_B1 = cov[B == 1,]
all(cov_B1$individualIdentifier == colnames(z.B1$z))
test_avez(z.B1$z, B1.genes, cov_B1)

# B3 test
cov_B3 = cov[B == 3,]
all(cov_B1$individualIdentifier == colnames(z.B1$z))
test_avez(z.B3$z, B3.genes, cov_B3)

```


# Figure 4A and 4B
* Figure 4A: Individual z-score heatmap for ROSMAP C0 subjects for Cluster #1.
* Figure 4B: Individual z-score heatmap for ROSMAP C0 subjects for Cluster #4.

```{r figure-4ab}

# prepare data
zC0 = z.C0$z
zC0 = zC0[order(clust.C0$clustA),]
zC0.genes = zC0[rownames(zC0) %in% C0_c4.genes,]

# color function
colFun1 = colorRamp2(c(-1, -0.5, 0, 0.5, 1), 
            c("#4575b4", "#74add1", "white", "#f46d43", "#d73027"))
colFun2 = colorRamp2(c(-4, -2, 0, 2, 4), 
            c("#0000FFFF", "#7C50FDFF", "#EEEEEEFF", "#FF6545FF", "#FF0000FF"))
clusters = c(1,4)

# Figure 4
panel_fun = function(index, nm) {
  if(index %in% which(sort(clust.C0$clustA) %in% clusters)){
    pushViewport(viewport())
    grid.rect(gp = gpar(fill = "white", col = "white"))
    grid.lines(c(0, 1, 1, 0), c(0, 0, 1, 1), gp = gpar(col = "#AAAAAA"), 
               default.units = "npc")
    hD = zC0
    topAnnot = HeatmapAnnotation(z = anno_barplot(colMeans(hD[index,]), 
                smooth = TRUE, axis_param = list(gp = gpar(fontsize = 5))),
                annotation_height = unit(0.7, "cm"))
    h = Heatmap(hD[index,], 
              name = "z-score",
              col = colFun2,
              show_row_names = FALSE,
              show_column_names = FALSE,
              cluster_rows = TRUE,
              show_heatmap_legend = FALSE,
              column_split = factor(z.C0$E, levels = c("E2", "E3", "E4")),
              cluster_column_slices = FALSE,
              column_title_gp =  gpar(col = "black", fontsize = 12),
              show_column_dend = TRUE,
              show_row_dend = TRUE,
              border = "#AAAAAA",
              row_dend_side = "right",
              height = unit(4.2, "cm"),
              column_dend_height = unit(0.7, "cm"),
              row_dend_width = unit(0.7, "cm"),
              width = unit(18, "cm"),
              top_annotation = topAnnot
              )
    draw(h, newpage = FALSE)
    popViewport()       
  }
}

# box for Figure 4
zoom_idx1 = which(rownames(zC0) %in% C0_c1.genes)
zoom_idx2 = which(rownames(zC0) %in% C0_c4.genes)
layer_fun = function(j, i, x, y, width, height, fill) {
  v = pindex(zC0, i, j)
  if(i %in% zoom_idx2) {
    grid.rect(gp = gpar(lwd = 2, col = "black"))
  }
  if(i %in% zoom_idx1) {
    grid.rect(gp = gpar(lwd = 2, col = "black"))
  }
}

anno = function(c, clust){
 return(anno_zoom(align_to = which(clust %in% c), 
                 which = "row", 
                 panel_fun = panel_fun, 
                 width = unit(21.5, "cm"), 
                 gap = unit(5, "cm"),
                 size = unit(7, "cm"),
                 link_width = unit(2, "cm"),
                 link_height = unit(5, "cm"),
                 link_gp = gpar(fill = "white", col = "#AAAAAA"), internal_line = FALSE)) 
}


# legend
lgd.h = Legend(col_fun = colFun1, title = "A: Average z-score", border = "black", 
               legend_width = unit(9.2, "cm"), 
               direction = "horizontal")

lgd.have = Legend(col_fun = colFun2, title = "B: z-score", border = "black", 
                  legend_width = unit(9.2, "cm"), 
                  direction = "horizontal")

pd = packLegend(lgd.h, lgd.have, 
                column_gap = unit(1, "cm"),
                direction = "horizontal")


# Figure 4A
pdf("../results/Figure4AB.cluster1.pdf", width = 11)
Heatmap(as.matrix(avez.C0[order(clust.C0$clustA), c("E2", "E3", "E4")]), 
        col = colFun1,
        cluster_columns = FALSE,
        cluster_rows = FALSE,
        show_row_names = FALSE,
        show_column_names = TRUE,
        right_annotation = rowAnnotation(foo = anno(1, sort(clust.C0$clustA))),
        row_split = sort(clust.C0$clustA), 
        column_names_side = "bottom",
        layer_fun = layer_fun,
        show_heatmap_legend = FALSE,
        width = unit(5, "cm"),
        column_names_rot = 0,
        column_names_centered = T,
        row_title_gp = gpar(fontsize = 11)
)
draw(pd, x = unit(18, "cm"), y = unit(0.95, "cm"))
grid.text("A", x = unit(0.5, "cm"), y = unit(17.2, "cm"), gp = gpar(fontsize=16))
grid.text("B", x = unit(8.5, "cm"), y = unit(17.2, "cm"), gp = gpar(fontsize=16))
dev.off()

# Figure 4B
pdf("../results/Figure4AB_cluster4.pdf", width = 11)
Heatmap(as.matrix(avez.C0[order(clust.C0$clustA), c("E2", "E3", "E4")]), 
        col = colFun1,
        cluster_columns = FALSE,
        cluster_rows = FALSE,
        show_row_names = FALSE,
        show_column_names = TRUE,
        right_annotation = rowAnnotation(foo = anno(4, sort(clust.C0$clustA))),
        row_split = sort(clust.C0$clustA), 
        column_names_side = "bottom",
        layer_fun = layer_fun,
        show_heatmap_legend = FALSE,
        width = unit(5, "cm"),
        column_names_rot = 0,
        column_names_centered = T,
        row_title_gp = gpar(fontsize = 11)
)
draw(pd, x = unit(18, "cm"), y = unit(0.95, "cm"))
grid.text("A", x = unit(0.5, "cm"), y = unit(17.2, "cm"), gp = gpar(fontsize=16))
grid.text("B", x = unit(8.5, "cm"), y = unit(17.2, "cm"), gp = gpar(fontsize=16))
dev.off()

```


# Figure 4C and 4D

Violin plots where each dot represents a gene.

```{r figure-4cd}

my_violin = function(genes){
        mDataZ.c = z.C0$z[genes, ]
        mDataZ.c.e2 = mDataZ.c[,z.C0$E == "E2"]
        mDataZ.c.e3 = mDataZ.c[,z.C0$E == "E3"]
        mDataZ.c.e4 = mDataZ.c[,z.C0$E == "E4"]

        my.dat = list(`2` = colMeans(mDataZ.c.e2), 
                      `3` = colMeans(mDataZ.c.e3), 
                      `4` = colMeans(mDataZ.c.e4))
        my.df = reshape2::melt(my.dat)
        my.df$L1 = as.numeric(my.df$L1)
        colnames(my.df) = c("z-score", "APOE")
        my.df$Genes = unlist(lapply(my.dat, function(x) names(x)))

        ggthemr('greyscale')
        colnames(my.df) = c("z-score", "APOE.num", "Genes")
        my.df$APOE = as.factor(my.df$APOE.num)
  
        to_swap = c("#62bba5", # E4
                    "#785d37", # E3
                    "#ffb84d") # E2
        
        # figure label
        APOE_label = c("\u03B52", "\u03B53", "\u03B54")
        names(APOE_label) = c("E2", "E3", "E4")
  
        to_swap = c("#62bba5", # E4
            "#785d37", # E3
            "#ffb84d") # E2
  
# violin plot
fig4c = ggplot(my.df, aes(x = APOE.num, y = `z-score`, color = APOE)) +
  geom_violin(fill = "white", trim = FALSE) +
  geom_quasirandom(alpha = 0.3, width = 0.1, dodge.width = 0.9, varwidth = TRUE) +
  scale_color_manual(values = c("#62bba5", "#785d37", "#ffb84d")) +
  labs(y = "Average Z-Score", x = "APOE Genotype") +
  theme(text = element_text(size = 14),
        legend.position="bottom", legend.direction = "horizontal",
        axis.title.x = element_text(face = "bold", size = 11),
        axis.title.y = element_text(face = "bold", size = 11),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        legend.title = element_text(face = "bold")) +
  stat_summary(fun.data=mean_cl_normal, geom="errorbar", width=0.2, position = position_dodge(0.9), color = "black")
        
        return(fig4c)
}

fig4c.c1 = my_violin(C0_c1.genes)
fig4c.c4 = my_violin(C0_c4.genes)

fig4c.c1.p = annotate_figure(fig4c.c1, fig.lab = "C",
                             fig.lab.size = 16)
fig4c.c4.p = annotate_figure(fig4c.c4, fig.lab = "D", 
                             fig.lab.size = 16)

ggsave("../results/Figure4C_0505.pdf", fig4c.c1.p, width = 5, height = 4, device=cairo_pdf)
ggsave("../results/Figure4D_0505.pdf", fig4c.c4.p, width = 5, height = 4, device=cairo_pdf)
```


# Figure 4E and 4F

E4 vs. E3 average Z-score line plot from C0 to C3. Note that the warnings are due to the Greek letters in the plot.

```{r figure-4ef}

# remove subjects which have no APOE or CERAD record
sel = (!is.na(cov$APOE)) & (!is.na(cov$C))
mData.2 = mData[, sel]

# z-score
means = rowMeans(mData.2)
sds = apply(mData.2, 1, sd)
zscore = function(x) return((x - means)/sds)
mDataZ = apply(mData.2, 2, zscore)
cov[, EC := paste(paste0("C", C), E, sep = ":")]
EC = cov$EC[sel]

# average z-score
aveZ = data.table(Genes = rownames(mDataZ),
                  C0E2 = rowMeans(mDataZ[, EC == "C0:E2"]),
                  C0E3 = rowMeans(mDataZ[, EC == "C0:E3"]),
                  C0E4 = rowMeans(mDataZ[, EC == "C0:E4"]),
                        
                  C1E2 = rowMeans(mDataZ[, EC == "C1:E2"]),
                  C1E3 = rowMeans(mDataZ[, EC == "C1:E3"]),
                  C1E4 = rowMeans(mDataZ[, EC == "C1:E4"]),
                        
                  C2E2 = rowMeans(mDataZ[, EC == "C2:E2"]),
                  C2E3 = rowMeans(mDataZ[, EC == "C2:E3"]),
                  C2E4 = rowMeans(mDataZ[, EC == "C2:E4"]),
                        
                  C3E2 = rowMeans(mDataZ[, EC == "C3:E2"]),
                  C3E3 = rowMeans(mDataZ[, EC == "C3:E3"]),
                  C3E4 = rowMeans(mDataZ[, EC == "C3:E4"]))
 
aveZ2 = melt(aveZ, id.vars = c("Genes"))
colnames(aveZ2) = c("Genes", "EC", "aveZ") 
aveZ2[, CERAD := gsub("E[234]$", "", aveZ2$EC)]
aveZ2[, APOE := gsub("C[0123]", "", aveZ2$EC)]

# colors
ggthemr("fresh")
to_swap = c("#62bba5", # E4
            "#785d37", # E3
            "#ffb84d") # E2

# figure label
APOE_label = c("\u03B52", "\u03B53", "\u03B54")
names(APOE_label) = c("E2", "E3", "E4")

# warnings because greek letters
fig4.f = ggplot(aveZ2[Genes %in% C0_c4.genes], 
                aes(x = CERAD, y = aveZ, color = APOE)) + 
  geom_point() +
  geom_line(aes(group = factor(Genes)),
            color = "black",
            alpha = 0.1) +
  ylab("Average z-score") +
  xlab("CERAD NP score") +
  facet_grid(cols = vars(APOE), labeller = as_labeller(APOE_label)) +
  scale_color_manual(values = rev(to_swap), labels = APOE_label)
fig4.f = annotate_figure(fig4.f, fig.lab = "F", fig.lab.size = 16)
ggsave("../results/Figure4F-astro-C4.pdf", fig4.f, width = 10, height = 5, device=cairo_pdf)

# warnings because greek letters
fig4.e = ggplot(aveZ2[Genes %in% C0_c1.genes], 
                aes(x = CERAD, y = aveZ, color = APOE)) + 
  geom_point() +
  geom_line(aes(group = factor(Genes)),
            color = "black",
            alpha = 0.1) +
  ylab("Average z-score") +
  xlab("CERAD NP score") +
  facet_grid(cols = vars(APOE), labeller = as_labeller(APOE_label)) +
  scale_color_manual(values = rev(to_swap), labels = APOE_label)
fig4.e = annotate_figure(fig4.e, fig.lab = "E", fig.lab.size = 16)
ggsave("../results/Figure4E-astro-C1.pdf", fig4.e, width = 10, height = 5, device=cairo_pdf)

```
