---
title: ROSMAP Astrocyte Analysis
author: Effect of *APOE* Alleles on the Glial Transcriptome in Normal Aging and Alzheimer’s Disease
output:
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
    toc: yes
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(eval = FALSE)
```


# Data Setup

```{r}
rm(list = ls())
require(limma)
require(ComplexHeatmap)
require(data.table)
require(pheatmap)
require(ggplot2)
require(gridExtra)
require(ggthemr)
require(SNFtool)
require(circlize)
require(ggpubr)

ds = "ROSMAP"
dtype = "FPKM"
expSet = readRDS("ROSMAP-24.Rds")

# load expression data 
mData = expSet$mData
geneIds = rownames(mData)
samples = colnames(mData)
```


# Astrocyte Analysis

```{r}
celltype = "astro"

# select astrocyte genes (n = 405) only
genes = readLines("Data/Astrocyte Genes.txt")

# select astrocyte genes that exist in ROSMAP
# after selection (n = 400)
annot = fread("Data/ENSEMBL GRCh38.p7.csv")
ensemblId_genes = unique(annot$EnsemblID[annot$GeneSymbol %in% genes])
ensemblIds = intersect(geneIds,ensemblId_genes)
annot2 = unique(annot[, .(EnsemblID, GeneSymbol, Description)])
annot = unique(annot[, .(EnsemblID, GeneSymbol)])
annot2 = annot2[EnsemblID %in% ensemblIds,]
annot = annot[EnsemblID %in% ensemblIds,]
length(ensemblIds)
```


# Pre-Processing

```{r}
# data selection
sel.gen = geneIds %in% ensemblIds 
mDataG = mData[sel.gen,]  # processed data
mDataRawG = mDataG   # raw data
geneIdsG = geneIds[sel.gen]
geneSymbols = annot$GeneSymbol[match(geneIdsG, annot$EnsemblID)]

# normalization
cat("Data Type: ", dtype, "\n")
mDataG = log2(mDataG + 0.05)
cat("Data Process: log2(FPKM + 0.05)\n")

C = sapply(colnames(mData)[colnames(mData)!="Gene"], 
            function(x) expSet$strat[["CERAD"]][match(x, SubjectID)][["CERAD"]])
B = sapply(colnames(mData)[colnames(mData)!="Gene"], 
            function(x) expSet$strat[["Braak"]][match(x, SubjectID)][["Braak"]])
E = sapply(colnames(mData)[colnames(mData)!="Gene"], 
           function(x) expSet$strat[["APOE"]][match(x, SubjectID)][["APOE"]])
```


# Run Models
* Model 1: `exp ~ APOE + CERAD | exp ~ APOE + Braak`
* Model 2 baseline: `C0E4-C0E3, C0E2-C0E3 | B1E4-B1E3, B1E2-B1E3`
* Model 3: `C3E4-C3E3, C3E2-C3E3 | B3E4-B3E3, B3E2-B3E3`

```{r}
source("Models.R")


# run Model 1
tT_CERAD = run_model1("CERAD", mDataG, C, E, annot, "ROSMAP", "astro")
tT_Braak = run_model1("Braak", mDataG, B, E, annot, "ROSMAP", "astro")

nrow(tT_CERAD[E4vsE3_P.Value < 0.05,])
nrow(tT_Braak[E4vsE3_P.Value < 0.05,])

intersect(tT_Braak[E4vsE3_P.Value < 0.05]$GeneSymbol,tT_CERAD[E4vsE3_P.Value < 0.05]$GeneSymbol)

length(tT_CERAD[E4vsE3_P.Value < 0.05 & E4vsE3_logFC > 0,]$GeneSymbol)
length(tT_CERAD[E4vsE3_P.Value < 0.05 & E4vsE3_logFC < 0,]$GeneSymbol)
length(tT_Braak[E4vsE3_P.Value < 0.05 & E4vsE3_logFC > 0,]$GeneSymbol)
length(tT_Braak[E4vsE3_P.Value < 0.05 & E4vsE3_logFC < 0,]$GeneSymbol)

TableS2.astro = tT_CERAD[, .(EnsemblID, GeneSymbol, E4vsE3_logFC, E4vsE3_P.Value, E2vsE3_logFC, E2vsE3_P.Value)]


# run Model 2
tT2_CERAD = run_model2("CERAD", mDataG, C, E, annot, "ROSMAP", "astro") 
tT2_Braak = run_model2("Braak", mDataG, B, E, annot, "ROSMAP", "astro")

nrow(tT2_CERAD[E4vsE3_P.Value < 0.05,])
nrow(tT2_Braak[E4vsE3_P.Value < 0.05,])

intersect(tT2_Braak[E4vsE3_P.Value < 0.05]$GeneSymbol,tT2_CERAD[E4vsE3_P.Value < 0.05]$GeneSymbol)

length(tT2_CERAD[E4vsE3_P.Value < 0.05 & E4vsE3_logFC > 0,]$GeneSymbol) 
length(tT2_CERAD[E4vsE3_P.Value < 0.05 & E4vsE3_logFC < 0,]$GeneSymbol)
length(tT2_Braak[E4vsE3_P.Value < 0.05 & E4vsE3_logFC > 0,]$GeneSymbol)
length(tT2_Braak[E4vsE3_P.Value < 0.05 & E4vsE3_logFC < 0,]$GeneSymbol)


# run Model 3
tT3_CERAD = run_model3("CERAD", mDataG, C, E, annot, "ROSMAP", "astro") 
tT3_Braak = run_model3("Braak", mDataG, B, E, annot, "ROSMAP", "astro")

nrow(tT3_CERAD[E4vsE3_P.Value < 0.05,])
nrow(tT3_Braak[E4vsE3_P.Value < 0.05,])

intersect(tT3_Braak[E4vsE3_P.Value < 0.05,]$GeneSymbol,tT3_CERAD[E4vsE3_P.Value < 0.05,]$GeneSymbol)

length(tT3_CERAD[E4vsE3_P.Value < 0.05 & E4vsE3_logFC > 0,]$GeneSymbol) 
length(tT3_CERAD[E4vsE3_P.Value < 0.05 & E4vsE3_logFC < 0,]$GeneSymbol)
length(tT3_Braak[E4vsE3_P.Value < 0.05 & E4vsE3_logFC > 0,]$GeneSymbol)
length(tT3_Braak[E4vsE3_P.Value < 0.05 & E4vsE3_logFC < 0,]$GeneSymbol)


# run Model 4
tT4_CERAD = run_model4("CERAD", mDataG, C, E, annot, "ROSMAP", "microglia") 

nrow(tT4_CERAD[E4vsE3_P.Value < 0.05,])

length(tT4_CERAD[E4vsE3_P.Value < 0.05 & E4vsE3_logFC > 0,]$GeneSymbol)
length(tT4_CERAD[E4vsE3_P.Value < 0.05 & E4vsE3_logFC < 0,]$GeneSymbol)

length(tT4_CERAD[E2vsE3_P.Value < 0.05 & E2vsE3_logFC > 0,]$GeneSymbol)
length(tT4_CERAD[E2vsE3_P.Value < 0.05 & E2vsE3_logFC < 0,]$GeneSymbol)


# run Model 5
tT5_CERAD = run_model5("CERAD", mDataG, C, E, annot, "ROSMAP", "microglia") 

nrow(tT5_CERAD[C1E4vsC1E3_P.Value < 0.05,])
nrow(tT5_CERAD[C2E4vsC2E3_P.Value < 0.05,])

tT2_CERAD2 = tT2_CERAD
tT3_CERAD2 = tT3_CERAD
colnames(tT3_CERAD2) = gsub("E4vsE3", "C3E4vsC3E3",colnames(tT3_CERAD2))
colnames(tT3_CERAD2) = gsub("E2vsE3", "C3E2vsC3E3",colnames(tT3_CERAD2))
colnames(tT2_CERAD2) = gsub("E4vsE3", "C0E4vsC0E3",colnames(tT2_CERAD2))
colnames(tT2_CERAD2) = gsub("E2vsE3", "C0E2vsC0E3",colnames(tT2_CERAD2))
```


# Z-Score Analysis and Spectral Clustering

```{r}
# z-scores for C0 subjects only
cSubject = 0

# remove the 3 low-expressed genes: ENSG00000124091, ENSG00000252798, ENSG00000204345
mDataG1 = mDataG[!rownames(mDataG) %in% c("ENSG00000124091","ENSG00000252798","ENSG00000204345"),]

# select only the C0 subjects and remove the subject with APOE NA
mDataG1 = mDataG1[, which(C == cSubject)]
E.C0 = E[which(C == cSubject)]

# z-score
means = rowMeans(mDataG1)
sds = apply(mDataG1, 1, sd)
zscore = function(x) return((x - means)/sds)
mDataZscore = apply(mDataG1, 2, zscore)
rownames(mDataZscore) = annot$GeneSymbol[match(rownames(mDataZscore), annot$EnsemblID)]

# check should be TRUE
cat("Is Z-score correct? ", all(mDataZscore[99,] == (mDataG1[99,] - means[99])/sds[99]), "\n")
```


# Average Z-Score

```{r}
# average z-score
z = data.table(Genes = rownames(mDataZscore),
               E2 = rowMeans(mDataZscore[, E.C0 == "E2"]),
               E3 = rowMeans(mDataZscore[, E.C0 == "E3"]),
               E4 = rowMeans(mDataZscore[, E.C0 == "E4"]))
```


# Spectral Clustering

```{r}
# spectral clustering
# astrocyte C0: k = 5; c = 4 & c = 1
k = 5
c = 4

require(SNFtool)
set.seed(9)

zMtx = as.matrix(z[, c("E2", "E3", "E4")])
dissMatrix = dist(zMtx)
rownames(zMtx) = z$Genes

# compute similarity matrix
simMatrix = 1-as.matrix(dissMatrix)/max(dissMatrix)
 
# uses code from the Similarity Network Fusion Paper
# where kVal is the number of clusters you would like to partition
clust = SNFtool::spectralClustering(simMatrix, K = k)
clustLab = as.factor(clust)
```


# Figure 4A and 4B

Individual z-score heatmap for ROSMAP C0 subjects.

```{r}
# prepare data
E.C0 = factor(E[C == "0"], levels = c("E2", "E3", "E4"))
mDataZscore.H = apply(mDataZscore, 2, function(x) return(ifelse(x > 4, 4, x)))
mDataZscore.H = apply(mDataZscore.H, 2, function(x) return(ifelse(x < -4, -4, x)))


# color function
colFun1 = colorRamp2(c(-1, -0.5, 0, 0.5, 1), c("#4575b4", "#74add1", "white", "#f46d43", "#d73027"))
colFun2 = colorRamp2(c(-4, -2, 0, 2, 4), c("#0000FFFF", "#7C50FDFF", "#EEEEEEFF", "#FF6545FF", "#FF0000FF"))

clust = as.factor(sort(clustLab))

# Figure 4
# individual z-score heatmap for ROSMAP C0 subjects
panel_fun = function(index, nm) {
  if(index %in% which(clust %in% c(1, 4))){
        pushViewport(viewport())
    grid.rect(gp = gpar(fill = "white", col = "white"))
    grid.lines(c(0, 1, 1, 0), c(0, 0, 1, 1), gp = gpar(col = "#AAAAAA"), default.units = "npc")
    hD = mDataZscore.H[order(clustLab),]
    topAnnot = HeatmapAnnotation(z = anno_barplot(colMeans(hD[index,]), smooth = TRUE, axis_param = list(gp = gpar(fontsize = 5))),
                               annotation_height = unit(0.7, "cm"))
    print(sort(rownames(hD[index,])))
    h = Heatmap(hD[index,], 
              name = "z-score",
              col = colFun2,
              show_row_names = FALSE,
              show_column_names = FALSE,
              cluster_rows = TRUE,
              show_heatmap_legend = FALSE,
              column_split = factor(E.C0, levels = c("E2", "E3", "E4")),
              cluster_column_slices = FALSE,
              column_title_gp =  gpar(col = "black", fontsize = 15),
              show_column_dend = TRUE,
              show_row_dend = TRUE,
              border = "#AAAAAA",
              row_dend_side = "right",
              height = unit(4.2, "cm"),
              column_dend_height = unit(0.7, "cm"),
              row_dend_width = unit(0.7, "cm"),
              width = unit(18, "cm"),
              top_annotation = topAnnot
              )
    draw(h, newpage = FALSE)
    popViewport()       
  }
}

# heatmap for cluster 4
which(sort(clustLab) == 4)

# box for Figure 4
layer_fun = function(j, i, x, y, width, height, fill) {
  v = pindex(zMtx[order(clustLab),], i, j)
  if(i %in% 255:307) {
    grid.rect(gp = gpar(lwd = 2, col = "black"))
  }
  if(i %in% 1:58) {
    grid.rect(gp = gpar(lwd = 2, col = "black"))
  }
}

anno = function(cluster){
 return(anno_zoom(align_to = which(clust %in% c(cluster)), 
                 which = "row", 
                 panel_fun = panel_fun, 
                 width = unit(21.5, "cm"), 
                 gap = unit(5, "cm"),
                 size = unit(7, "cm"),
                 link_width = unit(2, "cm"),
                 link_height = unit(5, "cm"),
                 link_gp = gpar(fill = "white", col = "#AAAAAA"), internal_line = FALSE)) 
}


# legend
lgd.h = Legend(col_fun = colFun1, title = "A: Ave z-score", border = "black", 
               legend_width = unit(9.2, "cm"), 
               direction = "horizontal")

lgd.have = Legend(col_fun = colFun2, title = "B: z-score", border = "black", 
                  legend_width = unit(9.2, "cm"), 
                  direction = "horizontal")

pd = packLegend(lgd.h, lgd.have, 
                column_gap = unit(1, "cm"),
                direction = "horizontal")


# Figure 4
# heatmap for average z-scores of cluster 1
pdf("../TablesFigures/Figure4-1.cluster1.pdf", width = 11)
Heatmap(zMtx[order(clustLab),], 
        col = colFun1,
        cluster_columns = FALSE,
        cluster_rows = FALSE,
        show_row_names = FALSE,
        show_column_names = TRUE,
        right_annotation = rowAnnotation(foo = anno(1)),
        row_split = sort(clustLab), 
        column_names_side = "bottom",
        layer_fun = layer_fun,
        show_heatmap_legend = FALSE,
        width = unit(5, "cm"),
        column_names_rot = 0,
        column_names_centered = T,
        row_title_gp = gpar(fontsize = 11)
)
draw(pd, x = unit(18, "cm"), y = unit(0.95, "cm"))
grid.text("A", x = unit(0.5, "cm"), y = unit(17.2, "cm"), gp = gpar(fontsize=20))
grid.text("B", x = unit(8.5, "cm"), y = unit(17.2, "cm"), gp = gpar(fontsize=20))
dev.off()

# Figure 4
# heatmap for average z-scores of cluster 4
pdf("../TablesFigures/Figure4-1.cluster4.pdf", width = 11)
Heatmap(zMtx[order(clustLab),], 
        col = colFun1,
        cluster_columns = FALSE,
        cluster_rows = FALSE,
        show_row_names = FALSE,
        show_column_names = TRUE,
        right_annotation = rowAnnotation(foo = anno(4)),
        row_split = sort(clustLab), 
        column_names_side = "bottom",
        layer_fun = layer_fun,
        show_heatmap_legend = FALSE,
        width = unit(5, "cm"),
        column_names_rot = 0,
        column_names_centered = T,
        row_title_gp = gpar(fontsize = 11)
)
draw(pd, x = unit(18, "cm"), y = unit(0.95, "cm"))
grid.text("A", x = unit(0.5, "cm"), y = unit(17.2, "cm"), gp = gpar(fontsize=20))
grid.text("B", x = unit(8.5, "cm"), y = unit(17.2, "cm"), gp = gpar(fontsize=20))
dev.off()
```


# Figure 4C and 4D

Violin plots where each dot represents a gene.


```{r}
my_violin = function(c, TableS2.astro){
        mDataZscore.c = mDataZscore[TableS2.astro[clustLab == c, GeneSymbol], ]
        mDataZscore.c.e2 = mDataZscore.c[,E.C0 == "E2"]
        mDataZscore.c.e3 = mDataZscore.c[,E.C0 == "E3"]
        mDataZscore.c.e4 = mDataZscore.c[,E.C0 == "E4"]

        my.dat = list(`2` = rowMeans(mDataZscore.c.e2), `3` = rowMeans(mDataZscore.c.e3), `4` = rowMeans(mDataZscore.c.e4))
        my.df = reshape2::melt(my.dat)
        my.df$L1 = as.numeric(my.df$L1)
        colnames(my.df) = c("z-score", "APOE")
        my.df$Genes = unlist(lapply(my.dat, function(x) names(x)))


        ggthemr('greyscale')
        colnames(my.df) = c("z-score", "APOE.num", "Genes")
        my.df$APOE = as.factor(my.df$APOE.num)
  
        to_swap = c("#62bba5", # E4
                    "#785d37", # E3
                    "#ffb84d") # E2
  
        # violin plot
        fig4c.1 = ggplot(my.df, aes(x=APOE.num, y=`z-score`, group=APOE, fill = APOE)) + 
          geom_violin(trim=FALSE) +
          scale_fill_manual(values = rev(to_swap)) +
          theme(legend.position = "none") +
          ylab("Ave z-score") +
          xlab("APOE") + 
        geom_point(size=1, 
                   colour="black", 
                   alpha = 0.2,
                   position=position_jitter(width=0.025)) +
        geom_line(aes(group=Genes), 
                  alpha = 0.1,
                  color = "black")
  
        # box plot
        fig4c.2 = ggplot(data = my.df, aes(x=APOE, y=`z-score`, group=APOE, fill = APOE)) +
        geom_boxplot(outlier.colour="black", outlier.shape=16,
               outlier.size=2, notch=FALSE) +
        scale_fill_manual(values = rev(to_swap)) +
        ylab("Ave z-score")

        return(list(fig4c.1, fig4c.2))
}

fig4c.c1 = my_violin(1, TableS2.astro)
fig4c.c4 = my_violin(4, TableS2.astro)

fig4c.c1.p = ggarrange(plotlist = fig4c.c1, ncol = 2)
fig4c.c4.p = ggarrange(plotlist = fig4c.c4, ncol = 2)
fig4c.c1.p = annotate_figure(fig4c.c1.p, fig.lab = "C",
                             fig.lab.size = 20)
fig4c.c4.p = annotate_figure(fig4c.c4.p, fig.lab = "D", 
                             fig.lab.size = 20)

ggsave("../TablesFigures/Figure4C.pdf", fig4c.c1.p, width = 5.5, height = 4)
ggsave("../TablesFigures/Figure4D.pdf", fig4c.c4.p, width = 5.5, height = 4)
```


# Figure 4E and 4F

E4 vs. E3 average Z-score line plot from C0 to C3.

```{r}
# cluster 4 genes
C4.genes = TableS2.astro[clustLab == 4, GeneSymbol]
C1.genes = TableS2.astro[clustLab == 1, GeneSymbol]


# remove the 3 low-expressed genes: ENSG00000124091, ENSG00000252798, ENSG00000204345
mDataG1 = mDataG[!rownames(mDataG) %in% c("ENSG00000124091","ENSG00000252798","ENSG00000204345"),]
  

# remove subjects which have no APOE or CERAD record
sel = (!is.na(E)) & (!is.na(C))
mDataG1 = mDataG1[, sel]

means = rowMeans(mDataG1)
sds = apply(mDataG1, 1, sd)
zscore = function(x) return((x - means)/sds)
mDataZscore = apply(mDataG1, 2, zscore)
  
# check should be TRUE
cat("Is Z-score correct? ", all(mDataZscore[99,] == (mDataG1[99,] - means[99])/sds[99]), "\n")

EC = paste(paste0("C", C[sel]), E[sel], sep = ":")
# average z-score
aveZ = data.table(Genes = annot$GeneSymbol[match(rownames(mDataZscore), annot$EnsemblID)],
                        
                        C0E2 = rowMeans(mDataZscore[, EC == "C0:E2"]),
                        C0E3 = rowMeans(mDataZscore[, EC == "C0:E3"]),
                        C0E4 = rowMeans(mDataZscore[, EC == "C0:E4"]),
                        
                        C1E2 = rowMeans(mDataZscore[, EC == "C1:E2"]),
                        C1E3 = rowMeans(mDataZscore[, EC == "C1:E3"]),
                        C1E4 = rowMeans(mDataZscore[, EC == "C1:E4"]),
                        
                        C2E2 = rowMeans(mDataZscore[, EC == "C2:E2"]),
                        C2E3 = rowMeans(mDataZscore[, EC == "C2:E3"]),
                        C2E4 = rowMeans(mDataZscore[, EC == "C2:E4"]),
                        
                        C3E2 = rowMeans(mDataZscore[, EC == "C3:E2"]),
                        C3E3 = rowMeans(mDataZscore[, EC == "C3:E3"]),
                        C3E4 = rowMeans(mDataZscore[, EC == "C3:E4"]))
 
ggthemr("fresh")
aveZ2 = melt(aveZ, id.vars = c("Genes"))
colnames(aveZ2) = c("Genes", "EC", "aveZ") 

aveZ2[, CERAD := gsub("E[234]$", "", aveZ2$EC)]
aveZ2[, APOE := gsub("C[0123]", "", aveZ2$EC)]

to_swap = c("#62bba5", # E4
            "#785d37", # E3
            "#ffb84d") # E2

fig4.f = ggplot(aveZ2[Genes %in% C4.genes], aes(x = CERAD, y = aveZ, color = APOE)) + 
  geom_point() +
  geom_line(aes(group = factor(Genes)),
            color = "black",
            alpha = 0.1) +
  ylab("AveZ") +
  xlab("CERAD NP score") +
  facet_grid(cols = vars(APOE)) +
  scale_color_manual(values = rev(to_swap))
fig4.f = annotate_figure(fig4.f, fig.lab = "F", fig.lab.size = 20)
ggsave("../TablesFigures/Figure4F-astro-C4.pdf", fig4.f, width = 10, height = 5)


fig4.e = ggplot(aveZ2[Genes %in% C1.genes], aes(x = CERAD, y = aveZ, color = APOE)) + 
  geom_point() +
  geom_line(aes(group = factor(Genes)),
            color = "black",
            alpha = 0.1) +
  ylab("AveZ") +
  xlab("CERAD NP score") +
  facet_grid(cols = vars(APOE))+
  scale_color_manual(values = rev(to_swap))
fig4.e = annotate_figure(fig4.e, fig.lab = "E", fig.lab.size = 20)
ggsave("../TablesFigures/Figure4E-astro-C1.pdf", fig4.e, width = 10, height = 5)
```